Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-14 optim.qmd
update the squintability calculation and misc

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

@@

```{r}
#| label: fig-projection
#| fig.height: 6
#| fig.width: 10
#| fig.cap: Best projections found by the JS opitmiser in each simulation, sorted by the index value. The projection pursuit problem is to detect the 6D sine-wave structure using the spline index. Most simulations have detected a clear sine-wave pattern, while the last six show a straight line with clustering at the ends.. In this simulation, the success rate is 0.88 (44/50).
#set.seed(123456)
#sin1000 <- spinebil::sinData(6, 1000) %>% scale() %>% as_tibble()
#colnames(sin1000) <- paste0("V", 1:6)
#remotes::install_github("uschiLaa/spinebil")
#load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sin1000.rda"))
#set.seed(123456)
#sin1000 <- spinebil::sinData(6, 1000) %>% scale() %>% as_tibble()
#colnames(sin1000) <- paste0("V", 1:6)

proj_dt <- map_dfr(sim_sine_6d_spline_best |> pull(basis), 
              ~as_tibble(as.matrix(sin1000) %*% .x), .id = "sim") |> 
  mutate(sim = as.numeric(sim)) |> 
  left_join(sim_sine_6d_spline_best |> select(sim, index_val), by = "sim")

@@

*Note: should we print out the 52 cases investigated somewhere, say
appendix, since it is not full grid simulation (otherwise for the
sine-wave problem there will be 6 indexes x 6 configurations = 36 cases
for 6-D data and 2 indexes \[MIC & TIC\] x 6 configurations = 12 cases
for 8-D data)? We only do a selection of 28 of them to save time/
computational resources.*

@@

*Note: Shall I repeat the scaled logistic function with our
parameterisation here? It can be handy to interpret theta1 - theta4 but
they should've already been explained in the squintability section.*

@tbl-smoothness-squintability presents the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for each case. The column
"smooth" is used as the smoothness measure and the column "squint" is
calculated as $\frac{\theta_1 \theta_2 \theta_3}{4}$ as squintability.
The low squint value of MIC and TIC index is due to its convex shape of
the index value against the projection distance, as shown in
@fig-idx-proj-dist. Comparing to other concave shapes, the maximum first
derivative happens at a smaller projection distance ($\theta_2$),
requiring the optimiser to get closer to the optimal to see a
significant change in the index value, hence a small squintability
measure.

*Note: the stringy squintability doesn't make sense*

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (including variance, range, smoothness, and nugget) and scaled logistic function ($\theta_1$ to $\theta_4$) for the pipe-finding and sine-wave finding problems.  The "smooth" column and "squint" column, calculated as $\frac{\theta_1 \theta_2 \theta_3}{4}$, represent the smoothness and squintability measures. 
dt <- tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness)
dt |> 
  knitr::kable(digits = 3, booktabs = T, escape = FALSE, row.names = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(3, 7, 12, 13), border_right = TRUE) |> 
  kable_styling(latex_options = "hold_position", font_size = 7, full_width = FALSE)
```

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first 7 rows of the datasets processed for modelling. 
sim_df |> 
  dplyr::select(index, d, smoothness, squintability, n_jellies, max_tries, 
                -I_max_max, P_J_hat, time) |> 
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 3, col.names = c("Index", "D", "Smoothness", "Squintability", 
                                  "Num. of jellyfish", "Max. Num. of tries", "Prop. of success", "Time"), 
        booktabs = T) |> 
  kable_styling(latex_options = "hold_position", font_size = 7)
```

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with the JS optimiser success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>  
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 30, 1, 0),
         P_J_hat = ifelse(index == "stringy", 0, P_J_hat))

mod1 <- glm(P_J_hat ~ smoothness + squintability + d + long_time + n_jellies + max_tries, data = sim_df2, family = binomial)
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness", "Squintability", "dimension (d)", 
                  "long time", "number of jellyfish", "maximum number of tries")) |>
  kable(digits = 3) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE)
```

@@

# Additional plot:

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values against projection distance for the 12 pipe/sine-wave finding problem after the binning procedure during the estimation of the squintability measure. The index value is averaged at each bin width of 0.005 and the `TIC` index is scaled to 0-1 for comparison. When finding the sine-wave structure using the `MIC` and `TIC` index, a convex curved is observed, as opposed to the pipe-finding problem or the sine-wave finding problem with the `dcor2d_2`, `loess2d`, and `splines2d` index. When finding the sine-wave with the stringy index, the index shows an instantaneous jump to the optimum when close to the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = idx, group = index)) +
  geom_line() +
  facet_nested_wrap(~index + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value")

>>>>

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#install the latest version of ferrn from: https://github.com/huizezhang-sherry/ferrn
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

TODO: add equation reference to main text line 705

@@

```{r}
#| label: fig-projection
#| fig.height: 6
#| fig.width: 10
#| fig.cap: Best projections found by the JS opitmiser in each simulation, sorted by the index value. The projection pursuit problem is to detect the 6D sine-wave structure using the spline index. Most simulations have detected a clear sine-wave pattern, while the last six show a straight line with clustering at the ends.. In this simulation, the success rate is 0.88 (44/50).
proj_dt <- map_dfr(sim_sine_6d_spline_best |> pull(basis), 
              ~as_tibble(as.matrix(sine1000) %*% .x), .id = "sim") |> 
  mutate(sim = as.numeric(sim)) |> 
  left_join(sim_sine_6d_spline_best |> select(sim, index_val), by = "sim")

@@

@@

@tbl-smoothness-squintability presents the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for each case. The column
"smooth" is used as the smoothness measure and the column "squint" is
calculated as [TODO: equation reference] as squintability.
The low squint value of MIC and TIC index is due to its convex shape of
the index value against the projection distance, as shown in
@fig-idx-proj-dist. Comparing to other concave shapes, the maximum first
derivative happens at a smaller projection distance ($\theta_2$),
requiring the optimiser to get closer to the optimal to see a
significant change in the index value, hence a small squintability
measure.

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values versus projection distance for the 12 pipe/sine-wave finding problem, after the binning procedure for calculating the squintability measure. The index values, averaged at bin width of 0.005, are scaled from 0 to 1 for comparison (holes, TIC, and stringy). The `MIC` and `TIC` index curves are convex while others are concave. The stringy curve shows an instantaneous jump to the optimum when aproaching the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = idx, group = index)) +
  geom_line() +
  facet_nested_wrap(~index + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value")

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (including variance, range, smoothness, and nugget) and scaled logistic function ($\theta_1$ to $\theta_4$) for the pipe-finding and sine-wave finding problems.  The column "smooth" and "squint" represent the smoothness and squintability measures. 
dt <- tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor2d", index)) 
dt |> 
  knitr::kable(
    digits = 3, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      "variance", "range", "smooth", "nugget",
      "$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", "squint"),
    booktabs = T, escape = FALSE, row.names = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(4, 8, 12, 13), border_right = TRUE) |> 
  kable_styling(font_size = 7)
```

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first 7 rows of the datasets processed for modelling. 
sim_df |> 
  dplyr::select(index, d, smoothness, squintability, n_jellies, max_tries, 
                -I_max, P_J, time) |> 
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 3,
        col.names = c("Index", "D", "Smoothness", "Squintability", 
                      "Num. of jellyfish", "Max. Num. of tries", 
                      "Prop. of success", "Time"), 
        booktabs = T) |> 
  kable_styling(latex_options = "hold_position", font_size = 7)
```

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with the JS optimiser success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>  
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 30, 1, 0),
         P_J = ifelse(index == "stringy", 0, P_J))

mod1 <- glm(P_J ~ smoothness + squintability + d + long_time + n_jellies + max_tries, data = sim_df2, family = binomial)
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness", "Squintability", "dimension (d)", 
                  "long time", "number of jellyfish", "maximum number of tries")) |>
  kable(digits = 3) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  kable_styling(font_size = 7)
```

@@
